Efficient Computation of Extended Surface Sources

### 2022-10-22 repro/papers/GEM22/submitR1/paper.tex
initial revision commit

\section{Introduction}
Full Waveform Inversion (FWI) can be described in terms of 
\begin{enumerate}
\item a linear wave operator $L[{\bf c}]$, depending on a vector of
  space-dependent coefficients ${\bf c}$ and acting on causal vector wavefields $\bu$ vanishing in negative time:
\begin{equation}
\label{eqn:init}
\bu \equiv 0, t \ll 0; 
\end{equation}
\item a trace sampling operator $P$ acting on wavefields and producing data traces;
\item and a (vector) source function (of space and time) $\bff$ representing energy input to the system. 
\end{enumerate}
The basic FWI problem is: given data $d$, find ${\bf c}$ so that 
\begin{equation}
\label{eqn:fwi}
P\bu \approx d \mbox{ and } L[\bf{c}]\bu = \bff.
\end{equation}
In this formulation, the source function $\bff$ may be given, or
to be determined subject to some constraints.

%The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time. In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there). Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
A simple nonlinear least squares formulation is:
\begin{equation}
\label{eqn:ols}
\mbox{choose } {\bf c} \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2.
\end{equation}
Practical optimization formulations typically augment the objective in
\ref{eqn:ols} by additive penalties or other constraints.

As is well-known, local optimization methods are the only feasible
approach given the dimensions of a typical instance of \ref{eqn:fwi},
and those have a tendency to stall due to ``cycle-skipping''. See for
example \cite{VirieuxOperto:09} and many references cited there. Source
extension is one approach to avoiding this problem. It consists in
imposing the wave equation as a soft as opposed to hard constraint, by
allowing the source field $\bff$ to have more degrees of freedom than
is permitted by a faithful model model of the seismic experiment, and
constraining these additional degrees of freedom by means of an
additive quadratic penalty modifying the problem \ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The operator $A$ penalizes deviation from known (or assumed)
characteristics of the source function - its null space consists of
feasible (or ``physical'') source models.

\cite{HuangNammourSymesDollizal:SEG19} present an overview of the
literature on source extension methods, describing a variety of
methods to add degrees of freedom to physical source model. The present paper
concerns {\em surface source extension}: physical sources are
presumed to be concentrated at points $\bx_s$ in space, whereas their extended
counterparts are permitted to spread energy over surfaces containing
the physical source locations. A simple choice for the penalty
operator $A$ is then multiplication by the distance $|\bx-\bx_s|$ to the physical
source location:
\begin{equation}
  \label{eqn:penop}
  (A\bff)(\bx,t) = |\bx-\bx_s|\bff(\bx,t)
\end{equation}
I shall use this choice of penalty operator whenever a specific choice
is necessary in the development of the theory below.

@@

The penalty operator $A$ defined in \ref{eqn:penop} is linear, so the source
subproblem is a linear least squares problem. Under some additional
assumptions to be described below, I shall show how to construct an
accurate approximate solution operator for problem
\ref{eqn:esis}. This approximate solution operator may be used to
accelerate Krylov space methods for the solution of the surface source
subproblem \ref{eqn:esis}. Numerical examples suggest the
effectiveness of this acceleration.

I will fully describe a preconditioner for a special
case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
acoustic field, $L[{\bf c}]$ is the wave operator of linear
acoustodynamics, the spatial positions of traces extracted by $P$ lie
on a depth plane $z=z_r$, and the positions at which the extended
source $\bff$ is nonzero lie on another, parallel, depth plane $z=z_s$. 
%That is, sources take the form of a combination of constitutive law
%defect and normal force,
%\begin{equation}
%  \label{eqn:surfsrc}
%  \bff(x,y,z,t) = h_s(x,y,t)\delta(z-z_s)(1,0,0,0)^T + f_s(x,y,z,t)(0,0,0,1)^T
%\end{equation}
%and the sampling operator $P$ is defined by
%\begin{equation}
%  \label{eqn:surfsam}
%  P\bu(x,y,t) =
%\end{equation}
%supported on $z=z_s$, and similar
%surface vertical loads.
This
``crosswell'' configuration simplifies the analysis underlying the
construction of approximate solutions for the source subproblem
\ref{eqn:esis}. It is only one of many transmission configurations for
which similar developments are possible. Perhaps the most important
alternative example is the diving wave configuration, which plays a
central role in contemporary FWI.

The preconditioner construction is very similar to the time-reversal
method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
norm in the domain and range spaces of the modeling operator. In this
case, the modfied norms are weighted $L^2$, and the weight operators
map pressure to corresponding surface source on the source and
receiver planes. This pressure-to-source map is closely related to the ``hyperbolic
Dirichlet-to-Neumann'' operator that plays a prominent role in
photoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
very similar preconditioner for Least Squares Migration, also for its
subsurface offset extension \cite[]{HouSymes:16}, motivated by
\cite{tenKroode:12}. These constructions also involve the
Dirichlet-to-Neumann operator. This concept also turns up in hidden
form in the work of Yu Zhang and collaborators on true amplitude
migration
\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.

The obvious computation of the pressure-to-source map - prescribe the
pressure, solve the wave equation with this boundary condition, read
off the equivalent source - suffers from intrinsic numerical
inaccuracy. I suggest an alternative computationally feasible
approach, via economical short-distance wave propagation. Since the
map is symmetric only in an approximate, asymptotic sense, it must be
symmetrized for use as a Krylov preconditioner. I describe a
symmetrization procedure that requires no further wave computations
beyond those necessary to compute the action of the operator itself.

The discussion in this paper is formal and incomplete, in the sense
that some important mathematical underpinnings are only
sketched. I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
it mapped square integrable surface sources to square integrable sampled
data. This is not true in full generality: while the surface source
problem has distribution solutions, they are not generally square
integrable (finite acoustic field energy). Even if the solutions have
finite energy, they do not in general have well-defined
restrictions to lower-dimensional sets. In other words, the action of
the sampling
operator $P$ on the space-time plane $z=z_r$ is not well-defined for arbitrary
finite-energy acoustic fields. Thus the modeling operator envisioned above
may not be well-defined.

This
phenomenon is related to the ill-posedness of wave equations as
evolution equations in spatial variables, an observation attributed to
Hadamard (see \cite{CourHil:62}, Chapter 6, section 17). Some constraint on the acoustic field,
beyond finite energy, is mandatory in any precise mathematical formulation the inverse problems
\ref{eqn:esi} and \ref{eqn:esis}. In fact, the natural constraint in
the ``crosswell'' geometry of this paper is that high-frequency energy
travel {\em only} along rays crossing the
surfaces $z=z_s, z=z_r$ transversally. That is, source functions on
$z=z_s$ generate waves with energy traveling along rays leaving the surface at a non-zero angle, and
energy arrives at the recording surface $z=z_r$ along rays making
non-zero angle with it. I will call sources, sampled data, and
acoustic fields with this property {\em
  downgoing} (even though the concept also encompasses {\em upcoming}
propagation). Note that the downgoing property restricts the behaviour of
acoustic fields near the source and receiver surfaces - what the
fields do elsewhere is their own business.

Several works have explored the mathematics of the
downgoing condition and its consequences in the context of the scalar second order wave equation, see for instance
\cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}. 
Elaboration of these mathematical details is beyond the scope of this
paper, which aims instead to explore the algorithmic
consequences of the mathematical structure implied by the downgoing
condition.

The next section defines the modeling operator $PL[{\bf c}]$, its
adjoint, and important specializations (pressure vs. normal velocity
sources and data). The sections to follow define the
source-to-pressure operator, construct an approximate inverse of the
modeling operator by time reversal (as suggested by work in
photoacoustic tomography), use the source-to-pressure operator to
express the approximate inverse as the modeling operator adjoint in
weighted norms (thus establishing that the modeling operator is {\em approximately
  unitary} in the sense of these norms), explain how to use this
construction to precondition Conjugate Gradient iteration, and
organize the preconditioning computation so as to involve only one
extra and relatively inexpensive
wave propagation calculation. The penultimate section displays simple
2D numerical examples of all of the key steps, culminating in a
comparison of straight vs. preconditioned CG iteration. The paper ends
with a brief discussion-and-conclusion section, reviewing what has
been accomplished and listing a few of the many questions left open.

@@

Most of what follows is valid for any space dimension $n >0$. The
coefficient vector $\bf{c}=(\kappa,\rho)$ is defined throughout space
$\bR^n$, the state vector $\bu$ throughout space-time
$\bR^{n+1}$. Whenever convenient for mathematical manipulations,
$n=3$: for instance, I will write $\bx=(x,y,z)^T$ for the spatial
coordinate vector, and refer to the third (vertical) coordinate of
particle velocity as $v_z$. Examples
later in this paper will use $n=2$ for computational convenience.

@@

The surface source extension replaces point sources on or near a
surface in $\bR^3$ with source functions confined to the same
surface. The simplest example of this extended geometry specifies a
plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
acoustic modeling, surface sources are combinations of constitutive law
defects and loads normal to the surface, localized on $z=z_s$. That
is, right-hand sides in the system $L\bu=\bff$ take the form
$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
takes the form
\begin{eqnarray}
\label{eqn:awepm}
\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} +
                                                f_s{\bf e} \delta(z-z_s),\nonumber \\
p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

@@

Extended forward modeling consists in solving \ref{eqn:awepm} and
sampling the solution components at receiver locations. For
simplicity, throughout this paper I will assume that the receivers are
located on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$ at
receiver depth $z_r>z_s$. The constructions to follow involve interchange
of the roles of $z_s$ and $z_r$ (that is, locating sources on $z=z_r$
and receivers at $z=z_s$), so rather than the sampling operator $P$ of
the introduction, I will denote by $P_s,P_r$ the sampling 
operators on $z=z_s$, $z=z_r$ respectively. In practice, sampling
occurs at a discrete array of points (trace locations) on these
surfaces, and over a zone of finite extent. In this theoretical
discussion, I will neglect both finite sample rate and extent, and
regard the data, for example $P_rp^+$, as continuously sampled and
extending over the entire plane $z=z_r$.

As explained in the Introduction, the downgoing constraint on the square-integrable
source functions $h_s, f_s$ is
essential both for finite energy solutions of the system
\ref{eqn:awepm} to exist, and for these solutions to have well-defined
traces on the receiver surface $z=z_r$. This constraint will be
assumed throughout, often tacitly.
For downgoing solutions of system \ref{eqn:awepm}, the key
components ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
in the open slab $z_s<z<z_r$ with well-defined limits at the boundary
planes, but may be discontinuous at the source plane
$z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be interchanged
in some of the constructions to come, and the corresponding solutions
may be discontinuous at $z=z_r$. Accordingly, interpret $P_s$, $P_r$
as the limit from right and left respectively: for $u=p^{\pm}$ or
$v^{\pm}_z$,
\begin{eqnarray}
  \label{eqn:defsamp}
  P_su(x,y,t) &=& \lim_{z \rightarrow z_s^+} u(x,y,z,t),\nonumber \\
  P_ru(x,y,t) &=& \lim_{z \rightarrow z_r^-} u(x,y,z,t).                  
\end{eqnarray}

The causal/anti-causal vector
modeling operators ${\cal S}^{\pm}_{z_s,z_r}$ are defined in terms of
the solutions $(p^{\pm},\bv^{\pm})$ of the systems \ref{eqn:awepm} by
\begin{equation}
  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)^T  = (P_rp^{\pm},P_r v_z^{\pm})^T,
  \label{eqn:fwd}
\end{equation}
The subscript signifies that sources are located on $z=z_s$, the
receivers on $z=z_r$. It is necessary to include this information in
the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
several locations will be needed in the discussion below.

\noindent {\bf Remark:} To connect with the formulation presented in
the introduction, note that for continuous $u$,
$P_su(x,y,t)=u(x,y,z_s,t)$, and therefore the adjoint of $P_s$ (in the
sense of distributions) is $P_s^Th(x,y,z,t) =
h(x,y,t)\delta(z-z_s)$. Write ${\cal P}_s = \mbox{diag }(P_s,P_s)$ and
similarly for ${\cal P}_r$. Then
\[
  {\cal S}^{+}_{z_s,z_r} = {\cal P}_r L^{-1}({\cal P}_s)^T,
\]
in which $L^{-1}$ is interpreted in the causal sense, and similarly
for ${\cal S}^{-}$. Sources confined to $z=z_s$ are precisely those
functions (distributions, really) output by ${\cal P}_s^T$, so the
problem statements \ref{eqn:esi} and \ref{eqn:esis} can be rewritten
in terms of ${\cal S}^+_{z_s,z_r}$, with $P$ identified with ${\cal P}_r$.

${\cal S}^{\pm}$ is not stably invertible: its columns are
approximately linearly dependent, as will be verified below. The
diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
its information, and it is in terms of these that a sensible inverse problem
is defined.

@@

>>>>

\section{Introduction}
Full Waveform Inversion (FWI) can be described in terms of 
\begin{enumerate}
\item a linear wave operator $L[{\bf c}]$, depending on a vector of
  space-dependent coefficients ${\bf c}$ and acting on causal vector wavefields $\bu$ vanishing in negative time:
\begin{equation}
\label{eqn:init}
\bu \equiv 0, t \ll 0; 
\end{equation}
\item a trace sampling operator $P$ acting on wavefields and producing data traces;
\item and a (vector) source function (of space and time) $\bff$ representing energy input to the system. 
\end{enumerate}
The basic FWI problem is: given data $d$, find ${\bf c}$ so that data
is fit and wave physics are honored, that is, 
\begin{equation}
\label{eqn:fwi}
P\bu \approx d \mbox{ and } L[\bf{c}]\bu = \bff.
\end{equation}
The source function $\bff$ may be given, or to be determined along
with $\bf{c}$.
%The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time. In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there). Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
The task \ref{eqn:fwi} may be cast as a nonlinear least squares problem: 
\begin{equation}
\label{eqn:ols}
\mbox{choose } {\bf c} \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2.
\end{equation}
Practical variants of the least squares problem \ref{eqn:ols}
typically augment the objective with additive penalties or other
constraints \cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}.

As is well-known, local optimization methods are the only feasible
approach given the dimensions of a typical instance of \ref{eqn:fwi},
and those have a tendency to stall at uninformative model estimates
due to ``cycle-skipping''. This phenomenon is often interpreted as the
capture of optimizing sequences at local minima of the nonlinear objective
function in \ref{eqn:ols} far from the global minimum, or at least not
possessing the characteristics of a useful solution. See for
example \cite{GauTarVir:86,VirieuxOperto:09,PladysBrossierLiMetivier:GEO21}.

Source extension is one approach to avoiding this``cycle-skipping''
obstacle. It consists in imposing the wave equation as a soft
constraint, allowing the source field $\bff$ to have more degrees of
freedom than is permitted by a faithful model model of the seismic
experiment, and constraining these additional degrees of freedom by
means of an additive quadratic penalty modifying the problem
\ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The linear operator $A$ penalizes deviation from known (or assumed)
characteristics of the source function - its null space consists of
feasible (or ``physical'') source models.
\cite{HuangNammourSymesDollizal:SEG19} present an overview of the
literature on source extension methods, describing a variety of
methods to add degrees of freedom to physical source model.

The present paper concerns {\em surface source extension}: physical
sources are presumed to be concentrated spatially at points $\bx_s$
in space, whereas their extended counterparts are permitted to spread
energy over a {\em source surface} containing the physical source spatial
locations. Similarly, the receiver spatial locations are confined to a
{\em receiver surface}. A simple choice for the penalty operator $A$ is then
multiplication by the distance $|\bx-\bx_s|$ to the physical source
location:
\begin{equation}
  \label{eqn:penop}
  (A\bff)(\bx,t) = |\bx-\bx_s|\bff(\bx,t)
\end{equation}
I shall use this choice of penalty operator whenever a specific choice
is necessary in the development of the theory below.

@@

The modeling operator $PL[{\bf c}]^{-1}$ and the penalty operator $A$ defined in \ref{eqn:penop} are linear, so the source
subproblem is a linear least squares problem. Under some additional
assumptions to be described below, I shall show how to construct an
accurate approximate solution operator for problem
\ref{eqn:esis}. This approximate solution operator may be used to
accelerate (``precondition'') Krylov space methods for the solution of the surface source
subproblem \ref{eqn:esis}. I will fully describe a preconditioner for a special
case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
acoustic field, $L[{\bf c}]$ is the wave operator of linear
acoustodynamics, and the source and receiver surfaces are horizontal
planes. Numerical examples in this setting suggest the
effectiveness of this acceleration.

I will use two 2D numerical models 
throughout to illustrate the theory. In both, horizontal lines serve as
source and receiver surfaces. The first is a ``crosswell'' or slab
configuration with an acoustic lens positioned between a deeps source
and a shallower line of 
receivers, resulting in markedly triplicated arrivals. The goal in
this first example is to construct a surface source that explains the data in a
homogeneous medium (that is, inversion in a wrong velocity, emulating
the early iterations of FWI based on extended modeling). The second is a layered
model in which velocity increases with depth, resulting the formation
of diving waves. This configuration simulates an
ocean-bottom node and a line of near-surface sources: the roles of
source and receiver are switched for computational convenience. In
this second example, the diving wave arrivals are isolated (by an
appropriate mute operator) and a
source constructed that explains them alone.

The next section defines the modeling operator $PL[{\bf c}]$, its
adjoint, and important specializations (pressure vs. normal velocity
sources and data). It also introduces the two 2D examples mentioned
above.  The following section constructs an approximate inverse of the
modeling operator by time reversal (as suggested by work in
photoacoustic tomography), and illustrates its efficacy in the two
examples. This construction requires extraction of
velocity data, or equivalently a surface source, from pressure data.
The subsequent sections describe this pressure-to-source operator,
express the approximate inverse as the modeling operator adjoint in
weighted norms (thus establishing that the modeling operator is {\em
  approximately unitary} in the sense of these norms), explain how to
use this construction to precondition Conjugate Gradient iteration,
and organize the preconditioning computation so as to involve only one
extra and relatively inexpensive wave propagation calculation. I use
the 2D examples to illustrate each of
these developments. The paper ends with a brief
discussion-and-conclusion section, reviewing what has been
accomplished and listing a few of the many questions left open.

%I will fully describe a preconditioner for a special
%case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
%acoustic field, $L[{\bf c}]$ is the wave operator of linear
%acoustodynamics, and the source and receiver surfaces are horizontal
%planes.
%spatial positions of traces extracted by $P$ lie
%on a depth plane $z=z_r$, and the positions at which the extended
%source $\bff$ is nonzero lie on another, parallel, depth plane $z=z_s$. 
%That is, sources take the form of a combination of constitutive law
%defect and normal force,
%\begin{equation}
%  \label{eqn:surfsrc}
%  \bff(x,y,z,t) = h_s(x,y,t)\delta(z-z_s)(1,0,0,0)^T + f_s(x,y,z,t)(0,0,0,1)^T
%\end{equation}
%and the sampling operator $P$ is defined by
%\begin{equation}
%  \label{eqn:surfsam}
%  P\bu(x,y,t) =
%\end{equation}
%supported on $z=z_s$, and similar
%surface vertical loads.
%This configuration simplifies the presentation of the
%approximate solution construction.

\section{Overview}

The preconditioner construction is closely related to the time
reversal method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09,Hristova:09}. A simplified mathematical
translation of this medical imaging task is to infer the initial
excess pressure distribution over a fluid-containing region at time
$t=0$ from measurements of the pressure on a surface enclosing the
fluid over a time interval $0 \le t \le t_{\rm max}$. This problem
clearly has strong similarities to, but also differences from, the
problem studied here. The time reversal method presumes that the
pressure field has returned to equilibrium (zero excess pressure), or close to it, at the
final time $t=t_{\rm max}$ Then the field can be (at least
approximately) viewed as the backwards-in-time initial boundary value
problem with zero final conditions at $t=t_{\rm max}$, and boundary
values given by the measurements. Evolving the field backwards in time
backwards to $t=0$ thus solves the problem. Except in special
circumstances, the pressure field never actually vanishes at finite
time, so the solution is approximate.

The seismic surface source extension proplem \ref{eqn:esis} differs in
several obvious ways from the photoacoustic setting. The measurement
or receiver surface does not surround the region of wave
propagation. It is not the initial pressure time-slice at $t=0$ that
is to be determined, but a time-extended source $\bff$ confined to a
surface. The penalty operator has no analogue in the basic
photoacoustic problem description. Finally, once these obstacles are
overcome, using the approximate solution operator to accelerate Krylov
iteration for solution of the optimization problem \ref{eqn:esis}
requires that the operator be identified as the adjoint of the
modeling operator with respect to suitable inner products in its domain
and range. In the pages to follow, I will address each of these issues.

First, reverse-time propagation can be localized via ray theory,
within high-frequency asymptotic approximation. This step requires
some assumptions about the ray fields: all rays responsible for significant
energy in the receiver data must arrive from the source surface, and must cross the source and receiver
surfaces transversally, that is, making non-zero angles. Also, the
rays must approach source and receiver surfaces from one side or the
other (``the inside''), so that locally the surfaces can be treated as boundaries of
propagation domains. With these assumptions, reverse-time propagation
of the receiver data
closely approximates the acoustic fields near the source surface. Next
observe that for a causal field with a surface pressure source (with a
continuous pressure field across the surface), the source is
proportional to the jump in the (particle) velocity field. Moreover,
to leading order in frequency, the velocity field switches sign at the
surface, so the jump is just twice the sampled value on the
surface. Thus reverse-time propagation and reading off the velocity
field on the source surface yields
a source that reproduces the pressure data, within an asymptotically small error.
%Preconditioning amounts to a change of
%norm in the domain and range spaces of the modeling operator. In this
%case, the modfied norms are weighted $L^2$, and the weight operators
%map pressure to corresponding surface source on the source and
%receiver planes.

This is the essence of the approximate solution of the problem
\ref{eqn:esis} for penalty weight $\alpha=0$. To see how Krylov
iteration might be accelerated, note that time-reverse propagation of
the receiver pressure values may be represented by a time-reverse
source propagation, with the source constructed from the velocity
field at the receiver surface. Moreover, the reverse-time
source-to-pressure propagation is the transpose (adjoint) of the
forward-time source-to-pressure propagation - this is a very simple
version of the adjoint state construction \cite[]{Plessix:06}. So the
approximate inverse is: conversion of pressure to source at the
receiver surface, followed by the adjoint of the modeling operator,
followed by the conversion of pressure to source at the source
surface. This sequence precisely describes the adjoint of the modeling
operator with respect to weighted norms in the spaces of source and
receiver data, with the weight operators being the
pressure-to-source operators. That this construction results in an
approximate inverse means that the modeling operator is approximately
unitary with respect to these weighted norms, which in turn implies
that the preconditioned conjugate gradient algorithm should converge rapidly.

This pressure-to-source map is closely related to the ``hyperbolic
Dirichlet-to-Neumann'' operator that plays a prominent role in
photoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
very similar preconditioner for Least Squares Migration, also for its
subsurface offset extension \cite[]{HouSymes:16}, motivated by
\cite{tenKroode:12}. These constructions also involve the
Dirichlet-to-Neumann operator. This concept also turns up in hidden
form in the work of Yu Zhang and collaborators on true amplitude
migration
\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.

FInally, the $\alpha=0$ case is not sufficient: the penalty operator
$A$ is an essential component of the nonlinear inverse propblem
\ref{eqn:esi}. As it turns out, $A$ commutes with the other operators
involved - approximately, but that is enough. Therefore the effect of
$A$ can be compensated with an easily-computed factor, whence the
preconditioned acceleration extended to the case $\alpha > 0$.

%The obvious computation of the pressure-to-source map - prescribe the
%pressure, solve the wave equation with this boundary condition, read
%off the equivalent source - suffers from intrinsic numerical
%inaccuracy. I suggest an alternative computationally feasible
%approach, via economical short-distance wave propagation. Since the
%map is symmetric only in an approximate, asymptotic sense, it must be
%symmetrized for use as a Krylov preconditioner. I describe a
%symmetrization procedure that requires no further wave computations
%beyond those necessary to compute the action of the operator itself.

The discussion in this paper is formal and incomplete, in the sense
that some important mathematical underpinnings are only
sketched. I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
it mapped square integrable surface sources to square integrable sampled
data. This is not true in full generality: while the surface source
problem has distribution solutions, they are not generally square
integrable (finite acoustic field energy). Even if the solutions have
finite energy, they do not in general have well-defined
restrictions to lower-dimensional sets. In other words, the action of
the sampling
operator $P$ on the receiver surface is not well-defined for arbitrary
finite-energy acoustic fields. Thus the modeling operator envisioned above
is not be well-defined, strictly speaking.

This phenomenon is related to the ill-posedness of wave equations as
evolution equations in spatial variables, an observation attributed to
Hadamard (see \cite{CourHil:62}, Chapter 6, section 17). A number of
authors have described precise forms of the ray conditions described
above, and shown how these conditions lead to the desired behaviour of
the modeling operator, that is, mapping of finite energy sources to
finite energy data, or equivalent properties
\cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}.
Elaboration of these mathematical details is beyond the scope of this
paper, which aims instead to explore the algorithmic consequences of
the mathematical structure implied by the nongrazing hypothesis.

%Some
%constraint on the acoustic field, beyond finite energy, is mandatory
%in any precise mathematical formulation the inverse problems
%\ref{eqn:esi} and \ref{eqn:esis}. I assume throughout that that
%high-frequency energy contributing to the measured data
%(output of the modeling operator) travels {\em only} along rays crossing the source
%and receiver surfaces transversally, that is, making non-zero angles
%with the tangent planes. I will call sources, sampled data, and
%acoustic fields with this property {\em non-grazing}. Note that the
%non-grazing property restricts the behaviour of acoustic fields near the
%source and receiver surfaces - what the fields do elsewhere is their
%own business.

%A further restriction turns out to be necessary: {\em all} of the rays
%associated with significan energy in the data must originate on
%the source surface and have the {\em incoming} property (essentially
%what is known in seismology as {\em downgoing}). This assumption uses
%an orientation of the source surface to distinguish inward and outward
%directions, and mandates that the rays in question all cross the
%source surface and propagate inwards in forward time. By symmetry,
%rays arriving at the receiver surface should propagate outward, with
%proper choice of orientation. These assumptions are also local at the
%source and receiver surfaces, and are often a natural consequence of
%the acquisition and ray geometry, as is the case in the two examples
%mentioned above.

@@

@@

Most of what follows is valid for any space dimension $n >0$. I will
describe the theory for $n=3$, write $\bx=(x,y,z)^T$ for the spatial
coordinate vector, and refer to the third (vertical) coordinate of
particle velocity $\bv$ as $v_z$. For computational convenience, the
examples are two-dimensional.

The surface source extension replaces point sources on or near a
surface in $\bR^3$ with source functions confined to the 
surface. The simplest example of this extended geometry specifies a
plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
acoustic modeling, surface sources are combinations of constitutive law
defects and loads normal to the surface, localized on $z=z_s$. That
is, right-hand sides in the system $L\bu=\bff$ take the form
$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
takes the form
\begin{eqnarray}
\label{eqn:awepm}
\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} +
                                                f_s{\bf e} \delta(z-z_s),\nonumber \\
p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

@@

Extended forward modeling consists in solving \ref{eqn:awepm} and
sampling the solution components at receiver locations. For
simplicity, throughout this paper I will assume that the receivers are
located on another spatial plane $\{(x,y,z,t): z=z_r\}$ at receiver
depth $z_r$. The constructions to follow involve interchange of the
roles of $z_s$ and $z_r$ (that is, locating sources on $z=z_r$ and
receivers at $z=z_s$), so both require sampling operators: $P_s,P_r$
are the sampling operators on $z=z_s$, $z=z_r$ respectively. In
practice, sampling necessarily occurs at a discrete array of points
(trace locations), and over a zone of finite extent. In this
theoretical discussion, I will neglect the finite sample rate, and
regard the data, for example $P_rp^+$, as continuously sampled. The
output samples are necessarily muted, that is, non-zero only over a space-time domain
of finite extent. This mute, and any tapering applied to the data
traces, are regarded as part of the sampling operators $P_s,P_r$.

The causal/anti-causal vector
modeling operators ${\cal S}^{\pm}_{z_s,z_r}$ are defined in terms of
the solutions $(p^{\pm},\bv^{\pm})$ of the systems \ref{eqn:awepm} by
\begin{equation}
  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)^T  = (P_rp^{\pm},P_r v_z^{\pm})^T,
  \label{eqn:fwd}
\end{equation}
The subscript signifies that sources are located on $z=z_s$,
receivers on $z=z_r$. It is necessary to include this information in
the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
several locations will be needed later in the discussion.

@@

The two examples mentioned in the introduction illustrate the setting
just described. The first example (Figure \ref{fig:olensrays0}) embeds
an acoustic (low-velocity) lens between a source surface at $z_s=3000$
(orange horizontal line) 
m and a receiver surface at $z_r=1000$ m (yellow horizontal line).  The data to be used in this
example results from a point source at $z_s=3500, x_s=3500$ m (orange
star). I have overlain the rays connecting this source point with the
data portion of the receiver surface: evidently all of the rays
involved cross the source and receiver surfaces transversally. Also,
regarding ``inwards'' as being ``upwards'' at the source surface,
``downwards'' at the receiver surface, it is clear that the data is
incoming at the source surface, outgoing at the receiver
surface.

\begin{figure}
  \includegraphics[width=\textwidth]{olensrays0.pdf}
  \caption{Bulk modulus with acoustic lens. Color scale unit is
    GPa. Orange horizontal line is source surface at depth $z_s= 3000$
    m, yellow horizontal
    line is receiver surface at depth $z_r=1000$ m. Point source location for data
    generation at $(3500, 3500)$ m indicated with star. Overlain with rays from point
    source to receiver surface.}
  \label{fig:olensrays0}
\end{figure}

%\ref{fig:drecplh0}, \ref{fig:dfwdplh0}
Figure \ref{fig:drecplh0} shows the gather $P_rp_{\rm pt}^{+}$, where
$p_{\rm pt}^{+}$ is causal field generated by the point source at
$(3500,3500)$ m with a trapezoidal bandpass filter wavelet having
significant energy between 1 and 12.5 Hz. The mute embedded in $P_r$
limits trace positions to 2000 m $ \le x_r \le $ 6000 m, and times to
1.2 s $ \le t \le $ 3 s.

Figure \ref{fig:dsrcphh0} displays an extended (pressure) source $h_s$, in the form of
traces in the spatial range 2000 m $ \le x_r \le $ 6000 m and time
range 0 s $\le t \le $ 2 s. The modeling operator output $S_{z_s,z_r}^+h_s$,
obtained from the causal solution $(p^+, \bv^+)$ of \ref{eqn:awepm}
with this choice of $h_s$ and $f_s=0,$ by application of the sampling
operator $P_r$ to $p^+$, are shown in Figure \ref{fig:dfwdplh0}.
